---
title: "uconn_modeling"
author: "Jack Weyer"
date: '2022-07-21'
output: pdf_document
---

```{r}
library(tidymodels)
library(MASS)
library(ggridges)
library(discrim)
library(vip)
library(ggrepel)
```

Load in data
```{r}
modeldf <- read.csv("UConnModelDF.csv")
modeldf <- modeldf %>% dplyr::select(-X)
modeldf <- modeldf %>% 
  mutate(success = success %>% as.factor(),
         Back = Back %>% as.factor(),
         player = player %>% as.factor())

```

Train/Test Split
```{r}
training <- modeldf %>% filter(year < 1903)
testing <- modeldf %>% filter(year == 1903)
```

##Cross validation
```{r}
set.seed(425)
folds <- vfold_cv(training, v = 10, repeats = 5)
keep_pred <- control_resamples(save_pred = TRUE, save_workflow = TRUE)
```

Every recipe will have airTime and distance with a combination of angle, Back, and/or Direction

##Create recipes
```{r}
recAll <- recipe(success ~ airTime + distance + Direction + angle, data = training) %>%
  step_dummy(all_nominal_predictors())

recAllBack <- recipe(success ~ airTime + distance + Back + angle, data = training) %>% 
  step_dummy(all_nominal_predictors())

recBack <- recipe(success ~ airTime + distance + Back, data = training) %>% 
  step_dummy(all_nominal_predictors())

recDir <- recipe(success ~ airTime + distance + Direction, data = training) %>% 
  step_dummy(all_nominal_predictors())

recAng <- recipe(success ~ airTime + distance + angle, data = training) %>% 
  step_dummy(all_nominal_predictors())

recNone <- recipe(success ~ airTime + distance, data = training) %>% 
  step_dummy(all_nominal_predictors())
```

##Logistic Regression
```{r}
lr_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```

```{r}
make_model <- function(spec, rec) {
  wf <- workflow() %>% 
    add_model(spec) %>% 
    add_recipe(rec)
  
  fitted <- wf %>% 
    fit_resamples(resamples = folds, control = keep_pred)
  
  collect_metrics(fitted)
}

TotalrecVec <- c("recAll", "recAllBack", "recBack", "recDir", "recAng", "recNone")
ModelList <- c()


run_models <- function(spec, recVec, ModelList) {
  for (i in recVec) {
    ModelList[[i]] <- make_model(spec, get(i))
  }
  return(ModelList)
}

LogRegModels <- run_models(lr_spec, TotalrecVec, ModelList)
LogRegModels
```
recBack wins the AUC contest.. (airTime + distance + Back)
accuracy 0.9725697
roc_auc	 0.9941544	

##Linear Discriminant Analysis
```{r}
lda_spec <- discrim_linear() %>% 
  set_mode("classification") %>% 
  set_engine("MASS")

LDAModels <- run_models(lda_spec, TotalrecVec, ModelList)
LDAModels
```
recNone wins the AUC contst.. (airTime + distance)
accuracy 0.9413377
roc_auc  0.9848817

##Quadratic Discriminant Analysis
```{r}
qda_spec <- discrim_quad() %>% 
  set_mode("classification") %>% 
  set_engine("MASS")

QDAModels <- run_models(qda_spec, TotalrecVec, ModelList)
QDAModels
```
recNone wins the AUC contst.. (airTime + distance)
accuracy 0.9673359
roc_auc  0.9923346

##Random Forest
```{r}
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) %>% 
  set_mode("classification") %>% 
  set_engine("ranger")

tune_wf <- workflow() %>% 
  add_recipe(recAll) %>% 
  add_model(tune_spec)

set.seed(98011)
tune_res <- tune_grid(
  tune_wf,
  resamples = folds,
  grid = 20
)

tune_res %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  dplyr::select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "AUC")
#mtry 2 to 3
#min_n 25 to 35

rf_grid <- grid_regular(
  mtry(range = c(2,3)),
  min_n(range = c(25,35)),
  levels=10
)

set.seed(509)
regular_res <- tune_grid(
  tune_wf,
  resamples = folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "AUC")

best_auc <- select_best(regular_res, "roc_auc")

final_rf <- finalize_model(
  tune_spec,
  best_auc
)

tree_prep <- prep(recAll)

final_rf %>%
  set_engine("ranger", importance = "permutation") %>%
  fit(success ~ .,
    data = juice(tree_prep)
  ) %>%
  vip(geom = "point")
```
```{r}
rf_spec <- rand_forest(mtry = 3, min_n = 32) %>% 
  set_engine("randomForest", importance = T) %>% 
  set_mode("classification")

rf_wf <- workflow() %>% 
  add_model(rf_spec) %>% 
  add_recipe(recAll)

rf_fit <- rf_wf %>% fit_resamples(resamples = folds, control = keep_pred)

collect_metrics(rf_fit)
```
accuracy 0.9697154	
roc_auc  0.9913012

Will try distance, airtime (recNone)
```{r}
tune_wf4 <- workflow() %>% 
  add_recipe(recNone) %>% 
  add_model(tune_spec)

set.seed(9)
tune_res4 <- tune_grid(
  tune_wf4,
  resamples = folds,
  grid = 20
)

tune_res4 %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  dplyr::select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
    values_to = "value",
    names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "AUC")
#mtry 1
#min_n 34 to 39

rf_grid4 <- grid_regular(
  min_n(range = c(34,39)),
  levels=6
)

tune_spec2 <- rand_forest(
  mtry = 1,
  trees = 1000,
  min_n = tune()
) %>% 
  set_mode("classification") %>% 
  set_engine("ranger")

tune_wf4 <- workflow() %>% 
  add_recipe(recNone) %>% 
  add_model(tune_spec2)

set.seed(1)
regular_res4 <- tune_grid(
  tune_wf4,
  resamples = folds,
  grid = rf_grid4
)

regular_res4 %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point() +
  labs(y = "AUC")
#mtry 1, min_n 38

best_auc4 <- select_best(regular_res4, "roc_auc")

final_rf4 <- finalize_model(
  tune_spec2,
  best_auc4
)

tree_prep4 <- prep(recNone)

final_rf4 %>%
  set_engine("ranger", importance = "permutation") %>%
  fit(success ~ .,
    data = juice(tree_prep4)
  ) %>%
  vip(geom = "point")
```

```{r}
rf_spec4 <- rand_forest(mtry = 1, min_n = 38) %>% 
  set_engine("randomForest", importance = T) %>% 
  set_mode("classification")

rf_wf4 <- workflow() %>% 
  add_model(rf_spec4) %>% 
  add_recipe(recNone)

rf_fit4 <- rf_wf4 %>% fit_resamples(resamples = folds, control = keep_pred)

RFmetrics <- collect_metrics(rf_fit4)
```
accuracy	0.9691202	50	0.0018111042	Preprocessor1_Model1
roc_auc	 0.9921972 .. best RF model


#Putting it all together
```{r}
modelviz <- data.frame(models = c("Log Reg", "LDA", "QDA", "RF"),
                       ROC_AUC = c(LogRegModels$recBack[["mean"]][2],
LDAModels$recNone[["mean"]][2], QDAModels$recNone[["mean"]][2],
RFmetrics$mean[2]))
```

```{r}
final_Model <- lr_spec %>% 
  fit(success ~ airTime + distance + Back, data = training)

augment(final_Model, testing) %>% conf_mat(truth = success, estimate = .pred_class)

augment(final_Model, testing) %>% conf_mat(truth = success, estimate = .pred_class) %>% autoplot()
```
97% accuracy on testing data!!!
```{r}
preds <- augment(final_Model, testing)

tidy(final_Model)
```
```{r}
preds <- preds %>% mutate(success = as.integer(success)-1)
preds <- preds %>% mutate(OAA = success - .pred_1)
OAAmetric <- preds %>% group_by(player) %>% 
  summarise(tot_OAA = sum(OAA),
            n = n()) %>% 
  arrange(desc(tot_OAA))

OAAmetric
```
Get Team A outfielders
```{r}
gameInfo <- read.csv("game_info.csv")
Agames <- gameInfo %>% filter(substr(game_str, 20,20) == 'A') %>% 
  filter(substr(game_str, 1, 4) == '1903')

Aplayers <- Agames %>% filter(top_bottom_inning == "Top") %>% pull(left_field) %>% unique()

Aplayers2 <- Agames %>% filter(top_bottom_inning == "Top") %>% pull(center_field) %>% unique()

Aplayers3 <- Agames %>% filter(top_bottom_inning == "Top") %>% pull(right_field) %>% unique()

Aplayers <- append(Aplayers, Aplayers2)
Aplayers <- append(Aplayers, Aplayers3) %>% unique()
```

Check if any Team A outfielders played against Team A
```{r}
Nonplayers <- Agames %>% filter(top_bottom_inning == "Bottom") %>% pull(left_field) %>% unique()

Nonplayers2 <- Agames %>% filter(top_bottom_inning == "Bottom") %>% pull(center_field) %>% unique()

Nonplayers3 <- Agames %>% filter(top_bottom_inning == "Bottom") %>% pull(right_field) %>% unique()

Nonplayers <- append(Nonplayers, Nonplayers2)
Nonplayers <- append(Nonplayers, Nonplayers3) %>% unique()

Nonplayers %in% Aplayers
```
Good. Can now assign teams confidently
```{r}
preds <- preds %>% mutate(Organization = ifelse(player %in% Aplayers, "A","Opponents"))
```

```{r}
byPlayer <- preds %>% group_by(player, Organization) %>% 
  summarise(tot_OAA = round(sum(OAA),1),
            n = n(),
            xOuts = sum(.pred_1),
            trueOuts = sum(success),
            OAAperOpportunity = round((tot_OAA/n),3)) %>% 
  arrange(desc(tot_OAA))

byPlayer %>% 
  ggplot(aes(x = xOuts, y = trueOuts, color = Organization)) + geom_point(size = 4, alpha = .7) +
  scale_color_manual(values = c("#003831", "purple")) +
  geom_label_repel(aes(
                label = ifelse(tot_OAA >= 1.5, as.character(player), "")),
                box.padding = 1.1,
                point.padding = 0.1,
                segment.color = "grey50") +
  geom_abline(color = "black") +
  theme_minimal() +
  labs(x = "Expected Outs (based on catch probabilities)",
       y = "Actual Outs",
       title = "1903 Outfielders by Organization",
       subtitle = "Line denotes average Outfielder. Above line = Above average") +
  theme(legend.position = c(.85,.3),
        legend.background = element_rect(fill = "white", color = "black"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 16))
```
```{r}
preds %>% 
  ggplot(aes(x = distance, y = airTime, color = .pred_1)) +
  geom_point() +
  scale_color_gradient2(midpoint = .5, high = "red", low = "blue", mid = "green") +
  theme_classic() +
  scale_x_continuous(n.breaks = 8) +
  scale_y_continuous(n.breaks = 7) +
  theme(axis.title.y = element_text(size=15),
        axis.title.x = element_text(size=13)) +
  labs(x = "Outfielder Distance at pitch to ball landing spot (feet)",
       y = "Hang Time (seconds)",
       color = "Catch\nProbability",
       title = "Predicted Outcomes on 1903 Data")
```
```{r}
preds %>% 
  mutate(Correct = ifelse(success == as.numeric(.pred_class)-1,1,0)) %>% 
  ggplot(aes(x = distance, y = airTime, color = as.factor(Correct))) +
  geom_point(alpha = 0.5) +
  theme_classic() +
  scale_x_continuous(n.breaks = 8) +
  scale_y_continuous(n.breaks = 7) +
  theme(axis.title.y = element_text(size=15),
        axis.title.x = element_text(size=13)) +
  labs(x = "Outfielder Distance at pitch to ball landing spot (feet)",
       y = "Hang Time (seconds)",
       color = "Predicted\ncorrectly?",
       title = "Predicted Outcomes on 1903 Data") +
  scale_color_manual(values = c("red", "gray"), labels = c("No", "Yes"))
```
```{r}
OAAmetric <- preds %>% group_by(player, Organization) %>% 
  summarise(`Season OAA` = round(sum(OAA),2),
            Opportunities = n()) %>% 
  arrange(desc(`Season OAA`))

OAAmetric
```

A or opponents better?
```{r}
preds %>% group_by(Organization) %>% 
summarise(tot_OAA = round(sum(OAA),1),
            n = n(),
            xOuts = sum(.pred_1),
            trueOuts = sum(success),
            OAAperOpportunity = round((tot_OAA/n),3)) %>% 
  arrange(desc(tot_OAA))
```

A by level
```{r}
Ateams <- preds %>% filter(Organization == "A") %>% 
  group_by(teamGame) %>% 
  summarise(tot_OAA = round(sum(OAA),1),
            n = n(),
            xOuts = sum(.pred_1),
            trueOuts = sum(success),
            OAAperOpportunity = round((tot_OAA/n),3)) %>% 
  arrange(desc(tot_OAA))

Ateams %>% ggplot(aes(x = teamGame, y = OAAperOpportunity)) + geom_bar(stat = "identity")
```
3 distinct levels

```{r}
TeamOpps <- preds %>%
  group_by(teamGame, Organization) %>% 
  summarise(tot_OAA = round(sum(OAA),1),
            n = n(),
            xOuts = sum(.pred_1),
            trueOuts = sum(success),
            OAAperOpportunity = round((tot_OAA/n),3)) %>% 
  arrange(desc(tot_OAA))

TeamOpps <- TeamOpps %>% mutate(trueTeam = paste(teamGame,Organization),
                                teamGame = paste(teamGame," Games"))



TeamOpps %>% 
  ggplot(aes(x = Organization, y = OAAperOpportunity, fill = Organization)) + geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#003831", "purple")) +
  facet_wrap(~teamGame) +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "red", size = 2) +
  labs(x = "",
       y = "OAA per Opportunity",
       title = "Organization A's Outfield Fly Ball Skill is topheavy",
       subtitle = "1903 season") +
  theme(legend.position = "none",
        axis.title.y = element_text(size = 17),
        strip.text.x = element_text(size = 14),
        axis.text.x = element_text(size = 12))
```

A1 > Opponents (+.046)
A2 < Opponents (-.009)
A3 < Opponents (-.011)

```{r}
TeamOpps %>% 
  group_by(Organization) %>% 
  summarise(Outs = sum(trueOuts),
            xOuts = sum(xOuts),
            OAA = Outs - xOuts,
            Opps = sum(n),
            OAApO = OAA/Opps)
```

Most improbable catch
```{r}
preds %>% filter(success==1) %>% filter(player == 2737) %>% 
  arrange(.pred_1)
```
Worst misses
```{r}
preds %>% filter(success==0) %>% arrange(desc(.pred_1))
```
Write results
```{r}
write.csv(byPlayer, "UConnResults.csv")
```

