---
title: "uconn_EDA"
author: "Jack Weyer"
date: '2022-07-20'
output: pdf_document
---
##Load Packages
```{r}
library(tidymodels)
library(MASS)
library(ggridges)
```

##Load Data
```{r}
teamInfo <- read.csv("team_info.csv")
ballPos <- read.csv("ball_pos.csv")
gameEvents <- read.csv("game_events.csv")
gameInfo <- read.csv("game_info.csv")
playerPos <- read.csv("player_pos.csv")
```

##Exploratory Data Analysis
```{r}
unique(teamInfo$team_id) #37 teams

gameInfo %>% 
  group_by(game_str, home_team) %>% 
  slice(1) %>% 
  count() %>% 
  group_by(home_team) %>% 
  summarise(homeGames = sum(n))

gameInfo %>% 
  group_by(game_str, away_team) %>% 
  slice(1) %>% 
  count() %>% 
  group_by(away_team) %>% 
  summarise(awayGames = sum(n))
```
A{1,2,3} and B are home teams and play lots of games. All other teams visit for 2-3 games.

Recorded Bounces Distribution
```{r}
bounces <-  gameEvents %>% 
  filter(event_code %in% c(9, 10, 16)) %>% 
  group_by(game_str) %>% 
  count()

fit <- fitdistr(bounces$n, densfun = "normal")

hist(bounces$n, prob=T, breaks=30)
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), add=T)
```
9 == ball deflection
10 === deflection off wall
16 == ball bounce
Very inconsistent wide distribution of recorded bounces across the 97 games. Several games (all Team B) with less than 10 "bounces."

```{r}
# add team and year variables to bounce df
bounces <- bounces %>% 
  mutate(year = as.numeric(substr(game_str,1,4))) %>% 
  mutate(team = ifelse(grepl("TeamB", game_str),1,0))
```

```{r}
bounces %>% 
  filter(team == 1) %>% #team B
  ggplot(aes(x = n, y = factor(year), fill = factor(year))) + geom_density_ridges() +
  theme_minimal() +
  labs(x = "Recorded Bounces + Deflections per Game",
       y = "",
       title = "Team B has inconsistent bounce tracking year to year") +
  theme(legend.position = "none") 
  
```
```{r}
bounces %>% 
  filter(team==0) %>% #Team A organization
ggplot(aes(x = n, y = factor(year), fill = factor(year))) + geom_density_ridges() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Recorded Bounces + Deflections per Game",
       y = "",
       title="Team A games have consistent (reliable) bounce data") 
```
```{r}
ballPos %>% 
  filter(game_str == "1900_01_TeamKJ_TeamB") %>% 
  filter(play_id == 14) %>%
ggplot(aes(x = timestamp, y = ball_position_z)) + geom_line() +
  geom_vline(xintercept = 367103, color = "green", lty = 2, size = 2) +
  geom_vline(xintercept = 370800, color = "red", lty = 2, size = 2) +
  #geom_vline(xintercept = 373704, color = "red", lty = 2, size = 2) +
  theme_classic() +
  labs(x = "Timestamp Duration",
       y = "Baseball Height (feet)",
       title = "Inconsistent Ball Tracking for Team B",
       subtitle = "Green = time of contact | Red = time that ball acquired")
```
Above is a play at Team B's field where it is clear the ball bounces twice but no bounce is recorded. This presents a huge problem with big data for to figure out if a fielder caught a ball in the air or after a bounce.

```{r}
# Filter to Team A's games
bounces <- bounces %>% 
  filter(team==0) %>% 
  mutate(home = substr(game_str, 20,21) %>% 
           as.factor())

bounces %>% 
  group_by(home, year) %>% 
  summarise(gamesPlayed=n())
```
Pretty consistent games played among the A organization. Limits outfield skill bias under assumption of three "levels"
```{r}
bounces %>% 
  ggplot(aes(x = n, y = factor(year), color = factor(year))) + geom_boxplot() + 
  geom_point() + 
  facet_wrap(~home) + 
  labs(title = "Bounce tracking appears consistent across year and level", y = "",
       x = "Bounces per game") +
  theme(legend.position = "none")
```
Is there a statistical difference?
```{r}
mod <- lm(data = bounces, n ~ year + home + year:home)
summary(mod)
```
Neither year nor "level" make a statistical impact on predicting recorded bounces per game among A games. 

```{r}
bounces <-  gameEvents %>% 
  filter(event_code %in% c(9, 10, 16)) %>% 
  group_by(game_str) %>% 
  count() %>% 
  mutate(year = as.numeric(substr(game_str, 1, 4))) %>% 
  mutate(team = ifelse(grepl("TeamB", game_str), 1, 0))

mod2 <- lm(data = bounces, n ~ year + team + year:team)
summary(mod2)
```
Team does make a statistical impact on bounces per game. We will exclude Team B games from this analysis.