---
title: "uconn_Cleaning"
author: "Jack Weyer"
date: '2022-07-20'
output: pdf_document
---

#Load Packages
```{r}
library(tidymodels)
library(MASS)
library(ggridges)
```

#Load Data
```{r}
teamInfo <- read.csv("team_info.csv")
ballPos <- read.csv("ball_pos.csv")
gameEvents <- read.csv("game_events.csv")
gameInfo <- read.csv("game_info.csv")
playerPos <- read.csv("player_pos.csv")
```

#Filter to Team A's games
```{r}
#As games
Agames <- gameInfo %>% 
  filter(home_team %in% c("TeamA1", "TeamA2", "TeamA3")) %>% 
  group_by(game_str) %>% 
  slice(1) %>% 
  pull(game_str)

AsGameInfo <- gameInfo %>% 
  filter(game_str %in% Agames)

AsGameEvents <- gameEvents %>% 
  filter(game_str %in% Agames)

AsBalls <- ballPos %>% 
  filter(game_str %in% Agames)

AsPlayers <- playerPos %>% 
  filter(game_str %in% Agames)

AsGameInfo <- AsGameInfo %>% 
  mutate(year = as.numeric(substr(game_str, 1, 4)))
```

Get game events (play_id) with a OF--2 (ball acquired), OF--7 (ball acquired unknown field position)

Other important player_position--event_code
 1--1 (pitcher pitch)
 2--2 (catcher catch)
 10--4 (ball in play)
 0--5 (end of play)
 x--3 throw
 x--9 player ball deflection
 255--10 deflection off wall
 255--16 ball bounce
 
#Filter our data to plays with outfielder "action"
```{r}
#plays where an outfielder is involved
Ofplays <- AsGameEvents %>% 
  filter(player_position %in% c(7,8,9)) %>% 
  dplyr::select(game_str, play_id) %>% 
  unique()

AsGameEvents <- inner_join(Ofplays, AsGameEvents)
AsBalls <- inner_join(Ofplays, AsBalls)
AsPlayers <- inner_join(Ofplays, AsPlayers)
```
#Get pitch "time" for each outfielder play
```{r}
pitches <- AsGameEvents %>% 
  filter(player_position == 1 & event_code == 1)

pitches$pitchTime <- pitches$timestamp

pitches <- pitches %>% dplyr::select(game_str, play_id, pitchTime)
AsGameEvents <- full_join(pitches, AsGameEvents)
```
#Get first "event" not including pitch or hit (OF ball acquired--2 or 7, catcher catch 2--2, OF deflection x--9, ball deflection 255--10, ball bounce 255-16)
```{r}
AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  group_by(event_code) %>% 
  summarise(n = n())
```
First event not including pitch and/or hit..
722 16's (Ball bounce)
691 2's (Ball acquired.. i.e. caught)
48 10's (deflection off wall)
14 9's (deflection) .. filter out deflections by non outfielders
11 5's (end of play) .. this is when end of play time = catch time
3 6's (pickoff) .. remove these

#Remove pickoffs
```{r}
pickoffs <- AsGameEvents %>% 
  filter(event_code == 6) %>% 
  dplyr::select(game_str, play_id)

AsGameEvents <- anti_join(AsGameEvents, pickoffs)
```
#Remove first deflections by non-outfielders
```{r}
DeflectedByNonOF <- AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% #first event after hit
  filter(event_code==9) %>% #deflection
  filter(!player_position %in% c(7,8,9)) %>% #not by an OF
  dplyr::select(game_str, play_id)

AsGameEvents <- anti_join(AsGameEvents, DeflectedByNonOF) #remove deflections by non OFs
```
#Remove non-batted ball plays (catcher errant throw to outfield)
```{r}
BattedBalls <- AsGameEvents %>% 
  filter(event_code==4) %>% 
  dplyr::select(game_str, play_id)

AsGameEvents <- inner_join(AsGameEvents, BattedBalls)
```

#On plays where the batted ball bounces before being picked up, filter to an OF touching the ball first
```{r}
BouncePlays <- AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  filter(event_code == 16) %>% #ball bounce
  dplyr::select(game_str, play_id)

bad_plays <- inner_join(BouncePlays, AsGameEvents) %>% filter(!event_code %in% c(1,4,16,10,5)) %>% 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  filter(!player_position %in% c(7,8,9)) %>% #non OFs
  dplyr::select(game_str, play_id) # these are plays that bounced but a non-outfilder was the first to deflect or acquire

AsGameEvents <- anti_join(AsGameEvents, bad_plays)
```

#Sometimes end of play is listed before a catch. Let's just remove 5's completely
```{r}
AsGameEvents <- AsGameEvents %>% 
  filter(event_code!=5)
```

Who touched first after hitting wall?
```{r}
WallPlays <- AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  filter(event_code==10) %>% #deflection off wall
  dplyr::select(game_str, play_id)


inner_join(WallPlays, AsGameEvents) %>%
  filter(!event_code %in% c(1,4,10, 16,5)) %>% 
  group_by(game_str, play_id) %>% 
  slice(1) %>%
  filter(!player_position %in% c(7,8,9))
  #Good! all "wall plays" were first picked up by an OF
```

Deflections need investigation.. were these all dropped?
```{r}
Deflections <- AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  filter(event_code==9) %>% 
  dplyr::select(game_str, play_id)

inner_join(Deflections, AsGameEvents)

#YES! checked the 10 manually. All dropped
```

The data has been properly subsetted to outfield opportunites for Team A's games.
```{r}
AsGameEvents %>% 
  filter(event_code != 1) %>% 
  filter(event_code != 4) %>% # not the pitch or hit 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  group_by(event_code) %>% 
  summarise(n = n())
```
712 16's (Ball bounce.. failure)
696 2's (Ball acquired.. success)
48 10's (deflection off wall.. failure)
10 9's (deflection and drop.. failure)

#Get ball landing (or catching) "time" and hang time for each play
```{r}
hm <- AsGameEvents
hm <- hm %>% 
  group_by(game_str, play_id) %>% 
  slice(3) %>% #after the pitch and ball in play events 
  mutate(endTime = timestamp) %>% #time of first event after contact
  mutate(airTime = endTime - pitchTime) %>% #hang time
  ungroup() %>% 
  dplyr::select(game_str, play_id, endTime, airTime)

AsGameEvents <- AsGameEvents %>% 
  inner_join(hm) #add hang time 
  
#Add year variable to differentiate training/testing values
AsGameEvents <- AsGameEvents %>% 
  mutate(year = as.numeric(substr(game_str,1,4)))

AsGameEvents %>% 
  filter(year != 1903) %>% 
  ggplot(aes(x = airTime)) + geom_density()
```
#Code success and failures
```{r}
hm <- AsGameEvents
hm <- hm %>% 
  group_by(game_str, play_id) %>% 
  slice(3) %>% 
  mutate(success = factor(if_else(event_code == 2,1,0))) %>% #ball acquired first is successful catch
  ungroup() %>% 
  dplyr::select(game_str, play_id, success)

AsGameEvents <- AsGameEvents %>% inner_join(hm)
```
#Hang time model effects
```{r}
AsGameEvents %>% 
  filter(year != 1903) %>% 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  mutate(airTime = airTime/1000) %>% 
  ggplot(aes(x = success, 
             y = airTime, 
             color = success)) + 
  geom_boxplot() + 
  geom_jitter(width = .25) +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Balls with more hang time are more likely to be caught",
       y = "Hang Time (seconds)",
       x = "",) +
  scale_x_discrete(labels = c("Dropped", "Caught")) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 18),
        axis.title.y = element_text(size = 15))
```
0-3 seconds: uncatchable
3-6 seconds: skill comes in to play
6+ seconds: can of corn

#Get each OF coordinates at the time of the pitch
```{r}
finaldf <- AsGameEvents %>% 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  dplyr::select(-at_bat, -player_position, -event_code)


# Each OF positions at the pitch
finaldf <- AsGameEvents %>% 
  group_by(game_str, play_id) %>% 
  slice(1) %>% 
  dplyr::select(game_str, play_id, timestamp) %>% 
  inner_join(playerPos) %>% 
  filter(player_position %in% c(7:9)) %>% 
  full_join(finaldf)
```
PROBLEM: Not all plays are included in game info (containing who is on the field)
```{r}
OFsByPlay <- AsGameInfo %>% 
  dplyr::select(game_str, play_per_game, left_field, center_field, right_field)

#plays where play_per_game matches 
have_player <- inner_join(OFsByPlay, finaldf)

#update. NAs need to be fixed
finaldf <- full_join(finaldf,have_player)

pbp_errors <- finaldf %>% 
  filter(is.na(left_field))


#51 errors
pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 1) %>% 
  ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)


new_players <- inner_join(OFsByPlay, pbp_errors)


have_player <- full_join(have_player,new_players, by = c("game_str", "play_id"))

##Add two to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 2) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add three to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 3) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add four to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 4) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add five to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 5) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add six to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 6) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add seven to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 7) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add eight to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 8) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add ten to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 10) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add eleven to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 11) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##Add 16 to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game + 16) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

##subtract 2 to play_per_game
pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))

pbp_errors <- pbp_errors %>% 
  mutate(play_per_game = play_per_game -2) %>% 
    ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))


#Left with 2 errors from the same game. Fix manually
pbp_errors <- pbp_errors %>% mutate(play_per_game = 99) %>% 
  ungroup() %>% 
  dplyr::select(game_str, play_per_game, play_id)

new_players <- inner_join(OFsByPlay, pbp_errors)

have_player <- full_join(have_player, new_players, by = c("game_str", "play_id"))

pbp_errors <- anti_join(finaldf, have_player, by = c("game_str", "play_id"))
```

One duplicate play.. De-duplicate that
```{r}
have_player %>% group_by(game_str, play_id) %>% count() %>% arrange(desc(n))

duplicatePlay <- have_player %>% filter(game_str == "1903_16_TeamNI_TeamA3", play_id == 75)

have_player <- have_player %>% filter(!(game_str == "1903_16_TeamNI_TeamA3" & play_id == 75)) %>% rbind(duplicatePlay[1:3,])
```

Add correct outfielders for each play
```{r}
CFs <- have_player %>% dplyr::select(game_str, play_id, contains("center")) %>% 
  mutate(across(contains("center"), ~paste(.), .names = "new_{.col}")) %>% 
  unite(CF, starts_with("new"), na.rm = T, sep = ",") %>% 
  dplyr::select(game_str, play_id, CF)

CFs$CF <- gsub('[,NA]','',CFs$CF)

RFs <- have_player %>% dplyr::select(game_str, play_id, contains("right")) %>% 
  mutate(across(contains("right"), ~paste(.), .names = "new_{.col}")) %>% 
  unite(RF, starts_with("new"), na.rm = T, sep = ",") %>% 
  dplyr::select(game_str, play_id, RF)

RFs$RF <- gsub('[,NA]','',RFs$RF)

LFs <- have_player %>% dplyr::select(game_str, play_id, contains("left")) %>% 
  mutate(across(contains("left"), ~paste(.), .names = "new_{.col}")) %>% 
  unite(LF, starts_with("new"), na.rm = T, sep = ",") %>% 
  dplyr::select(game_str, play_id, LF)

LFs$LF <- gsub('[,NA]','',LFs$LF)

finaldf <- cbind(finaldf, LFs)
finaldf <- cbind(finaldf, CFs)
finaldf <- cbind(finaldf, RFs)
```
#Get ball position at the catch
```{r}
finaldf <- finaldf %>% dplyr::select(game_str...1, play_id...2, player_position, field_x, field_y, pitchTime, play_per_game, endTime, airTime, year, success, LF, CF, RF)

finaldf <- finaldf %>% rename(game_str = game_str...1,
                              play_id = play_id...2)

BallEnds <- AsGameEvents %>% 
  filter(endTime == timestamp) %>% 
  inner_join(AsBalls)

#add columns for ball location at catch/drop
finaldf <- BallEnds %>% dplyr::select(game_str, play_id, ball_position_x, ball_position_y) %>% 
  inner_join(finaldf)
```
#Get distance of OFs at start of play and ball landing/catch spot
```{r}
# add distance
finaldf <- finaldf %>% rename(ball_end_x = ball_position_x,
                   ball_end_y = ball_position_y,
                   field_pitch_x = field_x,
                   field_pitch_y = field_y) %>% 
  mutate(distance = sqrt((ball_end_x - field_pitch_x)^2 + (ball_end_y - field_pitch_y)^2))
```

#Remove rows where a different outfielder made the catch
```{r}
# for balls caught by a different outfielder, we do not count that as an opportunity. 

catchInfo <- AsGameEvents %>% filter(timestamp==endTime) %>% 
  filter(success==1) %>% 
  dplyr::select(game_str, play_id, player_position) %>% 
  rename(caught = player_position)

finaldf <- full_join(finaldf, catchInfo)

finaldf <- finaldf %>% filter(is.na(caught) |
  caught == player_position)

finaldf <- finaldf %>% dplyr::select(-caught)
```

```{r}
finaldf %>% 
  filter(year != 1903) %>% 
  mutate(airTime = airTime / 1000) %>% 
  ggplot(aes(x = distance, 
             y = airTime, 
             color = factor(success))) + 
  geom_point(alpha = 0.4) +
  theme_classic() +
  labs(title="The relationship between hang time and distance is key",
       x = "Outfielder Distance at pitch to ball landing spot (feet)",
       y = "Hang Time (seconds)",
       color = "Caught?") +
  scale_color_manual(values = c("blue", "red"), labels = c("No", "Yes")) +
  theme(axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14))
```
#Get ball direction relative to outfielder facing directly towards home
```{r}
getOption("scipen")
options(scipen=999)
finaldf <- finaldf %>% 
  mutate(dotP = -1*field_pitch_x*(ball_end_x-field_pitch_x) + -1*field_pitch_y*(ball_end_y-field_pitch_y)) %>% 
  mutate(magnitudes = sqrt(field_pitch_x^2 + field_pitch_y^2) * sqrt(((ball_end_x-field_pitch_x)^2) + (ball_end_y-field_pitch_y)^2)) %>% 
  mutate(angle = acos(dotP / magnitudes)*180/pi)
```
Set directions to In, Back, or Side
```{r}
finaldf <- finaldf %>% 
  mutate(angle = round(angle, 2)) %>% 
  mutate(Direction = case_when(angle > 135 ~ "Back",
                               angle < 45 ~ "In",
                               TRUE ~ "Side"))
```

Dummy variable for fielder going back
```{r}
finaldf <- finaldf %>% mutate(Back = ifelse(Direction == "Back",1,0) %>% factor())
```

Cool viz
```{r}
library(GeomMLBStadiums)
finaldf %>% 
  filter(year != 1903) %>% 
  ggplot(aes(x = ball_end_x, y = ball_end_y, color = success)) +
  geom_point()+
  geom_spraychart(stadium_transform_coords = T, stadium_segments = "all", stadium_ids = "rangers") +
  theme_minimal() +
  labs(x = "feet",
       y = "feet",
       color = "Caught?",
       title = "1901-1902 Outfield Opportunity Locations") +
  scale_color_manual(values = c("blue", "red"), labels = c("No", "Yes")) 
```
Set the player to be the correct player (LF, CF, or RF)
```{r}
finaldf <- finaldf %>% mutate(player = case_when(player_position == 7 ~ LF,
                                      player_position == 8 ~ CF,
                                      TRUE ~ RF) %>% 
                     factor())
```
Assign home teams
```{r}
finaldf <- finaldf %>% mutate(teamGame = substr(game_str, 20, 21) %>% 
                                as.factor())
```

Make air time in seconds
```{r}
finaldf <- finaldf %>% 
  mutate(airTime = airTime/1000)
```

Drop unnecessary columns
```{r}
modeldf <- finaldf %>% 
  dplyr::select(-pitchTime, -play_per_game, -endTime, -LF, -CF, -RF, -dotP, -magnitudes)
```

Write out
```{r}
write.csv(modeldf, "UConnModelDF.csv")
```

